---
title: "def_prog_challenge"
output: html_document
date: "2024-09-18"
Team: "Butterfly"
---

## Basic ideas

*Defensive programming Principle 1. Before writing code, draft your program as a series of modular functions with high-level documentation, and develop tests for it.*

assumptions:

1.  the risk of influenza infection is 28% in each year of life

2.  the frequency of each circulating subtype could be inferred from the numbers of isolates sampled (primarily in hospitals) each year

let e_ij = probability of an individual with birth year i has his or her first IAV infection in calendar year j: e_ij = (1-a)\^(j-i) \* a where a is the risk of influenza infection in each year of life (0.28)

next, scaled e_ij by the fraction f_S of circulating IAVs belonging to subtype S (H1N1, H2N2, H3N2), so e_Sij = f_S\|j \* e_ij = f_S\|j \* (1-a)\^(j-i) \* a

finally, e_Si\|y is the probability that individuals born in i experienced their first infection with subtype S, given that y is current year (2017?). So we have: e_Si\|y = sum (e_Sij) from year i to year y

## codes

*Defensive programming Principle 2: Write clearly and cautiously. & Defensive programming Principle 3: Develop one function at a time, and test it before writing another*

Here I divided the code into a series of functions, each function only carry out one related calculation, and use a final function to wrap them up. Also I use Rmd and divided the whole program into code block with indexes, to make it more clear

*Principle 4: Document often & Principle 5. Refactor often*

We track and refactor our ideas and codes on Google doc and in this file

### read and modify input dataset

```{r cars}
library(reshape2) 
# read dataset
cases <- read.csv("data/subtype_counts.csv",stringsAsFactors = FALSE)
# sort columns by column names
cases <- cases[,order(colnames(cases))]
# mutate the dataframe referencing Year, so each row is (Year, Subtype, Count)
mut_cases <- melt(cases, id = c("Year")) 
```

### have a look on the dataset

```{r pressure, echo=FALSE}
# plot proportion of isolates sampled each year in the US
ggplot(mut_cases, aes(fill = variable, 
                      y = value, x = Year))+ 
  geom_bar(position = "fill", stat = "identity")+ 
  ggtitle("proportion of isolates sampled each year in the US")+ 
  labs(y="Proportion")+
  theme(plot.title = element_text(hjust = 0.5))+
  theme_cowplot()
```

### get first infection probability between time interval

```{r pressure, echo=FALSE}
# function to get first infection probability
get_first_infect_prob <- function(birth_year, year, infect_prob){
  # assert input is numbers
  stopifnot(is.numeric(birth_year) & is.numeric(year) & is.numeric(infect_prob))
  # assert input is positive
  stopifnot((birth_year)>0 & (year)>0 & (infect_prob)> 0)
  return((1-infect_prob) ^ (year-birth_year) * infect_prob)
}
get_first_infect_prob(1960, 1963, 0.28) # test code
```

### get subtype probability at a given year

```{r pressure, echo=FALSE}
get_subtype_prob <- function(year, subtype, df){
  # get year and subtype
  subtype_at_year <- df[df$Year == year & df$variable == subtype, ]
  # get total count of all subtypes at given year
  total_count <- df %>%
    group_by(Year) %>%
    summarise(Freq = sum(value))
  total_count <- total_count[total_count$Year == year, ]
  # return subtype proportion at given year
  return (subtype_at_year$value / total_count$Freq)
}
get_subtype_prob(1960, 'H2N2', mut_cases) # test code
```

### get subtype first infect probability at given year for given birth year

```{r pressure, echo=FALSE}
get_subtype_year_prob <- function(df, subtype,  birth_year, year, infect_prob){
  subtype_prob <- get_subtype_prob(year, subtype, df)
  infect_prob <- get_first_infect_prob(birth_year, year, infect_prob)
  # return subtype infect probability at given year for given birth year
  return(subtype_prob*infect_prob)
}
get_subtype_year_prob(mut_cases, 'H2N2', 1960, 1963, 0.28) # test code
```

### get probability of subtype first infection at a given year for given birth year

```{r pressure, echo=FALSE}
get_final_prob <- function(df, subtype,  birth_year, end_year, infect_prob){
  final_prob <- 0
  # from birth year to given end year, compute each year's probability and add them up
  for (x in birth_year: end_year){
    year_prob <- get_subtype_year_prob(df, subtype, birth_year, x, infect_prob)
    final_prob = final_prob+year_prob
  }
  return(final_prob)
}
get_final_prob(mut_cases, 'H2N2', 1960, 1968, 0.28) # test code
```

## wrap up

### get each year probability

For every birth year between 1960 and 1996, since the final year in the dataset is 2017, I will compute all the birth year to 2017 probability

```{r pressure, echo=FALSE}
# final function for computing probability of first infection for each subtypes

get_birthyear_prob <- function(df, end_year, infect_prob, birth_year_range, subtype_list ){
  final_df = NULL
  for (by in birth_year_range){
    # sapply function for each subtypes
      p <- sapply(subtype_list,get_final_prob,df=df, birth_year=by, 
                  end_year=end_year, infect_prob=infect_prob)
      p <- data.frame(as.list(p))
      final_df=do.call(rbind, list(final_df, p))
  }
  final_df <- cbind(final_df, birth_year=birth_year_range)
  return(final_df)
}

final <- get_birthyear_prob(mut_cases, 2017, 0.28,c(1960:1996), c('H1N1', 'H2N2', 'H3N2'))
final
```

### plot each year probability

```{r pressure, echo=FALSE}
# sort final df by column names
final <- final[,order(colnames(final))]
mut_final <- melt(final, id.vars="birth_year") 

# plot it out!
ggplot(mut_final, aes(fill = variable, 
                      y = value, x = birth_year))+ 
  geom_bar(position = "fill", stat = "identity")+ 
  ggtitle("Probability of first infection on 2017 for 1960-1996 borns in the US")+ 
  labs(y="probability")+
  theme(plot.title = element_text(hjust = 0.5))+
  theme_cowplot()

```

*Principle 6. When you run your code, save each “experiment” in a separate directory with a complete copy of the code repository and all parameters.*
